Supplementary Information

### 2020-08-13 doc/manuscript/MEE_manuscript.Rmd
a bit of work on manuscript

**[limitations in how traditional methods look at climate variables]**
- [subjectively selected variables (e.g., T, PPT, PDSI)]
- [month-by-month, reflective of fact that historical climate data products are often presented at this time scale. While useful for characterizing how monthly climate influences annual tree growth, month is an arbitrary unit of time and this approach does not necessarily identify the most influential time windows --often several months--over which climate shapes tree growth. (*note that we use the monthly time step here--don't set monthly resolution up as a problem we'll solve*)]
- [one variable at a time-- no additive or interactive effects (although VS-Lite starts to get there), which are expected and have been observed for tree-rings (Foster et al. 2016)]

**Traditional methods characterize only linear climate responses. This conflicts with biological expectations.**
Most biological rates—- from photosynthesis to animal metabolism—-display a unimodal relationship to temporal variation in temperature, wherein rates increase exponentially with temperature up to a point, typically reflective of the environment to which the organism is adapted / acclimatized, and decrease at higher temperatures (e.g., Slot and Winter 2017; Kumarathunge et al. 2019-cited in Helene's Tansley review). 
Similar relationships have also been observed for moisture availability, particularly in environments that are not strongly water-limited. 
In trees, responses of photosynthesis and respiration to environmental drivers are typically observed over time frames of seconds to days, and therefore do not directly reveal how annual tree growth and forest productivity respond to inter-annual variation in climate. 
The annual growth records of tree-rings allow can be used to study interannual variation, but the standard practice in dendrochronology has been to fit linear relationships (exception is VS-Lite, which fits plateau[@tolwinski-ward_bayesian_2013]), and we therefore know little about what, if any, nonlinearities occur in tree growth responses to interannual variation in climate.

**Traditional methods do not characterize the effect of tree size or its potential interactions with climate variables.** 
Tree size (most commonly diameter breast height, $DBH$) is among the most important variables affecting tree growth rate [@muller-landau_testing_2006; @foster_predicting_2016; REFS]. 
Radial growth increments (*i.e.*, tree-ring widths; $\Delta r$) may increase or decrease with tree size, often in a non-linear manner.
Following a "juvenile growth phase", which is typically removed in traditional dendrological analyses, $\Delta r$ may decline, particularly in open-grown conifers (??; DENDRO_REFS).
In contrast, in mesic closed-canopy forests, $\Delta r$ typically increases with tree diameter at breast height [$DBH$; @muller-landau_testing_2006; @anderson-teixeira_size-related_2015; REFS, DENDRO_REFS].
For dendrological studies aimed at deciphering climate signals, tree size is not typically a variable of interest, and its influence is removed through detrending (DENDRO_REFS).
While suitable for identifying climate signals (DENDRO_REFS), this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity. 
Although climate correlations can be transformed to climate sensitivity [*sensu* @charney_observed_2016]  and scaled to characterize the climate sensitivity of $ANPP_{stem}$ based on the scaling of $\Delta r$ with $DBH$ [@helcoski_growing_2019], they cannot be used to characterize known interactive effects of $DBH$ and climate on tree growth.
For example, larger trees tend to be more sensitive to drought [@bennett_larger_2015 ; McGregor et al. in revision; REFS].
Thus, to use tree-rings to predict $ANPP_{stem}$ for a forest where not every tree was cored, we need models that include tree size.

@@

@@

![**Figure 3 | Climate senstivity for all sites.** Columns include the top variables in the precipitation and temperature variable groups. For each species, relationships are plotted if included in top model. Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD. (IN-PREP FIGURE.) ](tables_figures/pre_temp_cld_groups.png)

@@

![**Figure 4 | Growth sensitivity to DBH: (a) $\Delta r$, (b) $BAI$, (c) $\Delta AGB$.** Only relationships included in the top model for each site are shown. It's tough to fit all the plots, so we may have to adjust this somehow.](tables_figures/DBH_responses.png)

@@

We get different climate sensitivity results with the different metrics of growth (Fig. 5?).
Preliminary review of some early results indicates that the "landscape" of climate effects over various time windows is generally similar across the three metrics of growth, but that the optimal time window or even the top climate variable in a group can shift.
(**See [ISSUE #40 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/40)**).

![**Figure 5 | Comparison of climwin output by growth metric:** Current example is PET at Lilly Dickey, selected mainly because data/model for other sites aren't yet stable. (I'm not sure if this is interesting enough to present in the main text. If not, we should have at least one example in the SI)](tables_figures/climwin_by_growth_metric.png)

>>>>

**To predict annual tree growth or forest productivity based on climate, a necessary first step is identification of the most important climate variables, the time frames over which they have the strongest influence over annual growth, and their additive or interactive effects.** 
Traditional dendrochronological methods identify climate signals by examining month-by-month correlations of growth index chronologies to select climate variables--typically T, PPT, and PDSI--over the current year and, typically, previous year. 
While appropriate for identifying the strongest signals for purposes such as historical climate reconstructions (e.g., DENDRO_REFS), this approach does not systematically examine the numerous potential multiple-month time windows to determine the single most influential window over which a given variable influences annual growth.
Needed is an objective approach to systematically evaluate the numerous potential climate variables and time windows  
Further, tree-ring studies generally do not consider additive or interactive effects of climate variables on annual growth [DENDRO_REFS; *but see VS-Lite* @foster_predicting_2016]. 
Such effects are expected based on observations that temperature and moisture interactively shape photosynthesis (REFS), tree growth [@foster_predicting_2016], and forest productivity [e.g., @taylor_temperature_2017; EXTRATROPICAL_AND_TREE_LEVEL_REFS; @banbury_morgan_global_nodate].

**Traditional methods characterize only linear climate responses. This conflicts with biological expectations.**
Most biological rates—- from photosynthesis to animal metabolism—-display a unimodal relationship to temporal variation in temperature, wherein rates increase exponentially with temperature up to a point, typically reflective of the environment to which the organism is adapted / acclimatized, and decrease at higher temperatures [e.g., @slot_high_2017] ([Kumarathunge et al. 2019](https://nph-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/nph.15668)). 
Similar relationships have also been observed for moisture availability, particularly in environments that are not strongly water-limited. 
In trees, responses of photosynthesis and respiration to environmental drivers are typically observed over time frames of seconds to days, and therefore do not directly reveal how annual tree growth and forest productivity respond to inter-annual variation in climate. 
The annual growth records of tree-rings allow can be used to study interannual variation, but the standard practice in dendrochronology has been to fit linear relationships (exception is VS-Lite, which fits plateau [@tolwinski-ward_bayesian_2013]), and we therefore know little about what, if any, nonlinearities occur in tree growth responses to interannual variation in climate.

**Traditional methods do not characterize the effect of tree size or its potential interactions with climate variables.** 
Tree size (most commonly diameter breast height, $DBH$) is among the most important variables affecting tree growth rate [@muller-landau_testing_2006; @foster_predicting_2016; REFS]. 
Radial growth increments (*i.e.*, tree-ring widths; $\Delta r$) may increase or decrease with tree size, often in a non-linear manner.
Following a "juvenile growth phase", which is typically removed in traditional dendrological analyses, $\Delta r$ may decline, particularly in open-grown conifers (??; DENDRO_REFS).
In contrast, in mesic closed-canopy forests, $\Delta r$ typically increases with tree diameter at breast height [$DBH$; @muller-landau_testing_2006; @anderson-teixeira_size-related_2015; REFS, DENDRO_REFS].
For dendrological studies aimed at deciphering climate signals, tree size is not typically a variable of interest, and its influence is removed through detrending (DENDRO_REFS).
While suitable for identifying climate signals (DENDRO_REFS), this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity. 
Although climate correlations can be transformed to climate sensitivity [*sensu* @charney_observed_2016]  and scaled to characterize the climate sensitivity of $ANPP_{stem}$ based on the scaling of $\Delta r$ with $DBH$ [@helcoski_growing_2019], they cannot be used to characterize known interactive effects of $DBH$ and climate on tree growth.
For example, larger trees tend to be more sensitive to drought [@bennett_larger_2015 ; @mcgregor_tree_nodate; REFS].
Thus, to use tree-rings to predict $ANPP_{stem}$ for a forest where not every tree was cored, we need models that include tree size.

@@

**This process identified different climate variables and time windows depending on the different metrics of growth (Fig. S2).**
Preliminary review of some early results indicates that the "landscape" of climate effects over various time windows is generally similar across the three metrics of growth, but that the optimal time window or even the top climate variable in a group can shift.
(**See [ISSUE #40 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/40)**).

@@

![**Figure 3 | Climate senstivity for all sites.** Columns include the top variables in the precipitation and temperature variable groups. For each species, relationships are plotted if included in top model. Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD. (IN-PREP FIGURE.) ](tables_figures/pre_temp_groups.png)

@@

![**Figure 4 | Growth sensitivity to DBH: (a) $\Delta r$, (b) $BAI$, (c) $\Delta AGB$.** Only relationships included in the top model for each site are shown.](tables_figures/DBH_responses.png)

@@
